---
title: "R Notebook"
output: html_notebook
---

```{r include=FALSE}
library(data.table)
library(linkcomm)

path <- "/Users/Carolina/Documents/GitHub/DegreeProject/" 
respath <- "/Users/Carolina/Documents/GitHub/DegreeProject/results/"

source(paste0(path, "code/myfunctions.R"))
```

```{r}
var.exp.lim <- 0.1

nSNPs <- 42052
nGenes <- 5720

snp.pval <- 0.01
snp.pval.nsign <- as.numeric(1e-5)

corr.pval <- 0.05/choose(nGenes,2)
```

```{r}
find.effects_TF <- fread(paste0(respath, "2020-01-27/findeffects_TF.gz"))
load(paste0(path,"results/2020-01-27/27_01_2020_linkedcommunities_TF.Rdata"))
```

```{r}
print(lc)
```


A data frame consisting of 2 columns; the first contains node names, and the second contains single community IDs for each node. All communities and their nodes are represented, but not necessarily all nodes.
```{r}
head(lc$nodeclusters)
```


A list of integer vectors containing the link IDs that belong to each community. Community IDs are the numerical position of the communities in the list.
```{r}
head(lc$clusters)
```


A data frame with 3 columns; the first two contain nodes that interact with each other, and the third is an integer vector of community IDs indicating community membership for each link.
```{r}
lc.edges <- data.table(lc$edges)
lc.edges
```



A named integer vector. Names are node names and integer values are the number of communities to which each node belongs.
```{r}
head(lc$numclusters)
```


A named integer vector. Names are community IDs and integer values indicate the number of nodes that belong in each community.
```{r}
head(lc$clustsizes)
tail(lc$clustsizes)
```

Nodes that belong to a lot of different communities will get the largest community centrality scores, whereas nodes that belong to overlapping, nested or few communities will get the lowest scores.
The 5 most central genes
```{r}
cc <- getCommunityCentrality(lc)
head(sort(cc, decreasing = TRUE))
```



```{r}
plot(lc, type = "members")
```
Numbers on the right - number of communities a gene belongs to
Numbers on the top - community(cluster) number



```{r}
gr <- graph_from_edgelist(as.matrix(lc.edges[cluster==39, .(node1, node2)]), directed=T)

plot(gr, layout=layout_with_graphopt)
```

get shared genes between two communities
```{r}
get.shared.nodes(lc, comms = c(1,7))
```

returns the meta-communities (clusters of community IDs)
```{r}
cr.20 <- getClusterRelatedness(lc, hcmethod = "ward", cutat = 20)
cr.5 <- getClusterRelatedness(lc, hcmethod = "ward", cutat = 5)
getClusterRelatedness(lc, hcmethod = "ward")
```

```{r}
cr.pdmax <- getClusterRelatedness(lc, hcmethod = "ward", cutat = lc$pdmax)
```



```{r}
plot(lc, type = "graph", shownodesin = 0, node.pies = T, vlabel=F, clusterids = cr.20[[1]])
plot(lc, type = "graph", shownodesin = 0, node.pies = T, vlabel=F, clusterids = cr.5[[1]], layout="spencer.circle")
```




```{r}
gr.meta <- graph_from_edgelist(as.matrix(lc.edges[cluster %in% cr.5[[1]], .(node1, node2)]), directed=T)

plot(gr.meta, vertex.label=NA)
```


```{r}
gr.meta2 <- graph_from_edgelist(as.matrix(lc.edges[, .(node1, node2)]), directed=T)

plot(gr.meta2, vertex.label=NA, vertex.color=lc.edges$cluster, layout=layout.fruchterman.reingold)
```



```{r}
cm <- getCommunityConnectedness(lc, conn = "modularity")
plot(lc, type = "commsumm", summary = "modularity")
```



collapse these larger numbers of communitites into smaller numbers of meta-communities
```{r}
mc <- meta.communities(lc, hcmethod = "ward", deepSplit = 0)
```

```{r}
plot(mc, type = "graph", vlabel = FALSE)
plot(lc, type = "graph", vlabel = FALSE)
```

