---
title: "R Notebook"
output: html_notebook
---

```{r include=FALSE}
library(data.table)
library(linkcomm)
path <- "/Users/Carolina/Documents/GitHub/DegreeProject/" 
respath <- "/Users/Carolina/Documents/GitHub/DegreeProject/results/"
source(paste0(path, "code/myfunctions.R"))
```

```{r}
find.effects_TF <- fread(paste0(respath, "2020-01-27/findeffects_TF.gz"))
load(paste0(path,"results/2020-01-27/27_01_2020_linkedcommunities_TF.Rdata"))
```


```{r}
print(lc)
```

# A data frame consisting of 2 columns; 
# the first contains node names, and the second contains single community IDs for each node. 
# head(lc$clusters)All communities and their nodes are represented, but not necessarily all nodes
```{r}
head(lc$nodeclusters)
```

# A list of integer vectors containing the link IDs that belong to each community. 
# each list has the members of that community
# Community IDs are the numerical position of the communities in the list.
```{r}
head(lc$clusters)
```


# A data frame with 3 columns; 
# the first two contain genes(nodes) that interact with each other
# the third is an integer vector of community IDs indicating community membership for each link.
```{r}
lc.edges <- data.table(lc$edges)
lc.edges
```

# Number of communities each gene(node) belongs to
```{r}
head(lc$numclusters)
```

# Names are community IDs
# integer values indicate the number of nodes that belong in each community.
```{r}
head(lc$clustsizes)
tail(lc$clustsizes)
```
# Nodes that belong to a lot of different communities will get the largest community centrality scores
# nodes that belong to overlapping, nested or few communities will get the lowest scores.
#The 5 most central genes
# cc <- getCommunityCentrality(lc)
#save(cc, file=paste0(respath, "2020-01-29/communitycentrality.Rdata"))
```{r}
load(paste0(respath, "2020-01-29/communitycentrality.Rdata"))
```

```{r}
head(sort(cc, decreasing = TRUE))
```
# plot genes that belong to the most communities

```{r}
plot(lc, type = "members")
```
# Numbers on the right - number of communities a gene belongs to
# Numbers on the top - community(cluster) number

# plot a specific cluster (or several )
```{r}
gr <- graph_from_edgelist(as.matrix(lc.edges[cluster %in% c(974, 988, 964) , .(node1, node2)]), directed=T)
plot(gr, layout=layout_with_graphopt, edge.arrow.size=.4, vertex.label=NA, edge.curved=.1)
```

```{r}
gr <- graph_from_edgelist(as.matrix(lc.edges[cluster==974 , .(node1, node2)]), directed=T)
plot(gr, layout=layout_with_graphopt, edge.arrow.size=.4, vertex.label=NA, edge.curved=.1)
```
# two communities that share genes
```{r}
gr <- graph_from_edgelist(as.matrix(lc.edges[cluster %in% c(1, 7) , .(node1, node2)]), directed=T)
plot(gr, layout=layout_with_graphopt, edge.arrow.size=.4, edge.curved=.1)
plot(lc, type = "graph", node.pies = T, vlabel=T, clusterids = c(1, 7))
```
# get shared genes between two communities
```{r}
get.shared.nodes(lc, comms = c(1,7))
```

# get nodes in certain communities
```{r}
getNodesIn(lc, clusterids = c(1,7))
```


Three measures of centrality include connectivity, closeness and betweenness. 
Connectivity - describes the number of links that a node has. 
Closeness centrality - the measure of distance from one node to all other nodes. 
Betweenness - the number of times a node is used as a bridge for other two nodes along the shortest paths. 
These centrality measures can determine how important a node in the graph and examine if the network has scale-free property

# Networks with high modularity have dense connections between the nodes within modules
# but sparse connections between nodes in different modules.
# Modularity is often used in optimization methods for detecting community structure in networks.

# measure of how relatively outwardly or inwardly connected a community is.
# names are community IDs and the numbers are community connectedness or modularity scores.
```{r}
#cconnect <- getCommunityConnectedness(lc)
#save(cconnect, file=paste0(respath, "2020-01-29/communityconnectedness.Rdata"))
load(paste0(respath, "2020-01-29/communityconnectedness.Rdata"))
```

```{r}
#cm <- getCommunityConnectedness(lc, conn = "modularity")
#save(cm, file=paste0(respath, "2020-01-29/communitymodularity.Rdata"))

load(paste0(respath, "2020-01-29/communitymodularity.Rdata"))
```

Table with cluster number and connectedness and modularity value
```{r}
con_mod <- data.table(cluster=1:length(cm),connectedness=cconnect, modularity=cm)
```

```{r}
con_mod_numnodes <- merge(con_mod,data.table(cluster=as.integer(names(lc$clustsizes)), numnodes=lc$clustsizes))
con_mod_numnodes <- con_mod_numnodes[order(-modularity)]
con_mod_numnodes
```


Plot the cluster with highest modularity
```{r}
plot(lc, type = "graph", node.pies = F, vlabel=T, vshape="circle" , vsize=3, ewidth=1, clusterids = con_mod_numnodes$cluster[1:10], layout="spencer.circle")
plot(lc, type = "graph", node.pies = F, vlabel=T, vshape="circle" , vsize=3, ewidth=1, clusterids = con_mod_numnodes$cluster[1:10])
plot(lc, type = "graph", node.pies = F, vlabel=T, vshape="circle" , vsize=3, ewidth=1, clusterids = 672, layout="spencer.circle")
```

```{r}
shared  <- get.shared.nodes(lc, comms = c(1969,971))
```

```{r}
find.effects_TF[geneA %in% shared | geneB %in% shared & geneA=="YLR245"]
```

```{r}
plot(lc, type = "commsumm", summary = "modularity")
```


# 10 communities with highest modularity - dense connections within community and sparce between nodes in different communities
```{r}
plot(lc, type = "commsumm", summary = "modularity", clusterids=as.integer(names(sort(cm, decreasing = T)[1:10])))

plot(lc, type = "graph", node.pies = T, vlabel=T, clusterids = as.integer(names(sort(cm, decreasing = T)[1:10])))

```

# 10 communities with highest connectivity - dense connections between communities
```{r}
plot(lc, type = "commsumm", summary = "connectedness", clusterids=as.integer(names(sort(cconnect, decreasing = T)[1:10])))

plot(lc, type = "graph", node.pies = T, vlabel=T, clusterids = as.integer(names(sort(cconnect, decreasing = T)[1:10])), layout="spencer.circle")

```

Table with number of clusters each gene belongs to and community centrality score
```{r}
genes_nclusters_centrality <- data.table(genes=names(cc), numclusters=lc$numclusters, centrality=cc)
genes_nclusters_centrality[order(-centrality)]
```


```{r}
get_clusters <- function(res){
  return(paste(unique(lc.edges[node1==res[[1]] | node2==res[[1]]]$cluster), collapse = ";"))
}
genes_nclusters_centrality2 <- data.table(genes_nclusters_centrality)
genes_nclusters_centrality2[, clustnum := mapply(get_clusters, genes)]
```

```{r}
plot(lc, type = "graph", node.pies = F, vlabel=F, vshape="circle" , vsize=5, ewidth=1, clusterids = as.integer(unlist(strsplit(genes_nclusters_centrality2[genes=="YOR342C"]$clustnum, ";"))))
```

```{r}
plot(lc, type = "graph", node.pies = F, vlabel=F, vshape="circle" , vsize=3, ewidth=1, clusterids = as.integer(unlist(strsplit(genes_nclusters_centrality2[genes=="YOR342C"]$clustnum, ";"))), layout="spencer.circle")
```

```{r}
plot(lc, type = "graph", node.pies = F, vlabel=T, vshape="circle" , vsize=3, ewidth=1, clusterids= c(991, 1445), layout="spencer.circle")
plot(lc, type = "graph", node.pies = F, vlabel=T, vshape="circle" , vsize=3, ewidth=1, clusterids = 3039)
```

```{r}
plot(lc, type = "graph", node.pies = F, vlabel=F, vshape="circle" , vsize=3, ewidth=1, shownodesin=200)
```

```{r}
plot(lc, type = "graph", node.pies = F, vlabel=F, vshape="circle" , vsize=5, ewidth=1, clusterids = as.integer(unlist(strsplit(genes_nclusters_centrality2[genes=="YMR029C"]$clustnum, ";"))), layout="spencer.circle")
```

```{r}
plot(lc, type = "graph", node.pies = F, vlabel=F, vshape="circle" , vsize=3, ewidth=1, clusterids =2938, layout="spencer.circle")
```

