---
title: "R Notebook"
output:
  html_notebook
---

### Set up
```{r eval=FALSE, include=TRUE}
library(data.table)
library(tidyr)
library(igraph)
library(dplyr)
library(parallel)

# path <- "/home/carolpb/DegreeProject/" # to use on Uppmax
# respath <- "/proj/snic2019-8-367/private/carol/results/"

respath <- "/Users/Carolina/Documents/GitHub/DegreeProject/results/"
path <- "/Users/Carolina/Documents/GitHub/DegreeProject/" #to use on my computer

source(paste0(path,"code/myfunctions.R"))

nSNPs <- 42052
nGenes <- 5720

snp.pval <- 0.05/(as.numeric(nGenes) * as.numeric(nSNPs))
snp.pval.nsign <- as.numeric(1e-5)

corr.pval <-  0.05/(nGenes*nGenes)

load(paste0(path, "results/2020-01-10/effectstable.Rdata"))
```


### Create table with cases where A->B is true and B->A is false
```{r}
find.effects <- effects_table.cor[cor.pval < corr.pval & cis.A ==T & cis.B==T & geneA!=geneB]

find.effects <- find.effects_fun(find.effects, snp.pval, snp.pval.nsign)

find.effects_TF.1 <- find.effects[find.effects$`A->B`==T & find.effects$`B->A`==F, .(geneA, geneB, eqtl.A, eqtl.B, `A->B`, `B->A`, eqtlA_geneB.pval)]
find.effects_TF.2 <- rbind(find.effects_TF.1, find.effects[find.effects$`A->B`==F & find.effects$`B->A`==T, .(geneA=geneB, geneB=geneA, eqtl.A=eqtl.B, eqtl.B=eqtl.A, `A->B`=`B->A`, `B->A`=`A->B`, eqtlA_geneB.pval=eqtlB_geneA.pval)])
find.effects_TF <- unique(find.effects_TF.2)
find.effects_TF
```
## Histogram of p-vals
```{r}
hist(find.effects_TF$eqtlA_geneB.pval)
hist(log10(find.effects_TF$eqtlA_geneB.pval))
```




## Gene pairs
Get table with how many times a gene pair appears - 
N==1 - these pairs only appear once
N==2 - these pairs appear twice
N==3 - these pairs appear three times 

This shows that most gene pairs only appear once
```{r}
find.effects_TF.numpairs <- find.effects_TF[(`A->B`==T & `B->A`==F), .N, by=.(geneA, geneB)]
numpairs.table <- data.table(N = unique(find.effects_TF.numpairs$N), numpairs =NA)
for (i in 1:nrow(numpairs.table)){
  numpairs.table$numpairs[i] <- nrow(find.effects_TF.numpairs[N==numpairs.table$N[i]])
}
numpairs.table
```

### plot the number of times a gene pair appears
```{r}
bp <- barplot(numpairs.table$numpairs, numpairs.table$N, names.arg=unique(find.effects_TF.numpairs$N), 
              width = 0.5, space=0.2, legend.text = F, ylim = c(0,max(numpairs.table$numpairs)+3000),
              main = "Number of times gene pairs appear \n (A->B = T and B->A = F)", xlab = "# times a gene pair appears", 
              ylab = "# of gene pairs")
text(bp,numpairs.table$numpairs, labels=numpairs.table$numpairs, cex=1, pos=3)
```

## Gene-eqtl pairs
```{r}
find.effects_TF.geneeqtl.A <- find.effects_TF[(`A->B`==T & `B->A`==F), .N, by=.(geneA, eqtl.A)]
find.effects_TF.geneeqtl.B <- find.effects_TF[(`A->B`==T & `B->A`==F), .N, by=.(geneB, eqtl.B)]

find.effects_TF.geneeqtl.A.plot <- find.effects_TF.geneeqtl.A %>% unite(gene_eqtlA, geneA, eqtl.A, sep = "__")
#find.effects_TF.geneeqtl.B.plot <- find.effects_TF.geneeqtl.B %>% unite(gene_eqtlB, geneB, eqtl.B, sep = "__")

plot(as.factor(find.effects_TF.geneeqtl.A.plot$gene_eqtlA), find.effects_TF.geneeqtl.A.plot$N, 
     xlab="gene-eqtl pairs", ylab="# times each pair appears", axes=FALSE, 
     main="Num times of times each gene-eqtl pair appears \n (A->B = T and B->A = F)")
Axis(side=2, labels=T)
```

### Find number of times a gene-eqtl pair appears (for geneA-eqtlA --> the "origin"/ the gene/eqtl that affects the other) 
```{r}
numpairs.table2 <- data.table(N = unique(find.effects_TF.geneeqtl.A.plot$N), numpairs =NA)
for (i in 1:nrow(numpairs.table2)){
  numpairs.table2$numpairs[i] <- nrow(find.effects_TF.geneeqtl.A[N==numpairs.table2$N[i]])
}
numpairs.table2 <- numpairs.table2[order(N)]
numpairs.table2
```

### Bar plot of table above
```{r}
# all values individually
bp2 <- barplot(numpairs.table2$numpairs, numpairs.table2$N, names.arg=unique(numpairs.table2$N), 
               width = 0.5, space=0.2, legend.text = F, ylim = c(0,max(numpairs.table2$numpairs)+100),
               main = "Number of times gene-eqtl pairs appear \n (A->B = T and B->A = F)", xlab = "# times a gene-eqtl pair appears", 
               ylab = "# gene-eqtl pairs", cex.names=0.8)
```

### Same, but grouping values 
```{r}
hist(find.effects_TF.geneeqtl.A.plot$N, main = "# gene-eqtl pairs", xlab = "# times gene-eqtl pairs appear",labels = T, ylim = c(0,max(find.effects_TF.geneeqtl.A.plot$N)+2000))
```
the majority of gene-eqtl pairs appear between 0-50 times

### Plot network
```{r}
plot_TF <- graph_from_edgelist(as.matrix(find.effects_TF[,.(geneA, geneB)]),directed=TRUE)
```

```{r}
V(plot_TF)$size <- 5
plot(plot_TF, layout=layout_with_fr,edge.arrow.size=.5, vertex.label=NA)
```

### Check how many groups of connected genes there are

```{r}
components(plot_TF)$no
```
if components(plot_TF)\$no > 1, there are several groups and those groups are not connected to each other  
if components(plot_TF)\$no == 1, all genes are connected

\n
table where you have each gene and which group it belongs to  
- not interesting if there's only one group
```{r}
memb <- components(plot_TF)$membership
nodes <- data.table(id = names(memb),
                    group_id = memb)
nodes <- nodes[order(nodes$id), ]
nodes
```



# Testing with different parameters
```{r}
sign_p <- c(1e-17, 1e-16, 1e-15, 1e-14, 1e-13, 1e-12, 1e-11, 1e-10, 1e-9, 1e-8, 1e-7, 1e-6, 1e-5, 1e-4, 1e-3, 1e-2)
non_sign_p <- c(1e-7, 1e-6, 1e-5, 1e-4, 1e-3, 1e-2)
cor_p <- c(1e-9, 1e-8, 1e-7, 1e-6, 1e-5, 1e-4, 1e-3, 1e-2)
params0 <- data.table(expand.grid(sign_p=sign_p, non_sign_p=non_sign_p, cor_p=cor_p))
```
Ran on diprotodon
```{r eval=FALSE, include=TRUE}
# cl = makeCluster(detectCores()-2, type="FORK")
# cl = makeCluster(4, type="FORK")
# res <- parApply(cl=cl,params0,1, testparams, effects_table.cor)
# stopCluster(cl)
```
the result is stored in res (list of tables)

```{r}
load(paste0(respath, "2020-01-20/20_01_2020_resparams.Rdata"))
```

## create summary table for number of unique gene pairs found and unique gene-eqtl pairs found (both for geneA (the one that has an effect) and for gene B)
one row per group of parameters
```{r}
cl = makeCluster(detectCores()-2, type="FORK")
res_table.temp <- parLapply(cl=cl, res, create_res_table)
stopCluster(cl)

res_table <- rbindlist(res_table.temp)
res_table <- res_table[order(sign.p, -nonsign.p, cor.p)]
res_table
```

## plot number of unique gene pairs found when using different parameters combinations
one panel for each corr.pval   
The line colors represent different values for the non-significant pval threshold
x-axis represents the significant p-val treshold

```{r}
par(mfrow=c(2,4))

cor.pvals <- unique(res_table$cor.p)
linetype <- c(1:length(unique(res_table$nonsign.p)))

for (p in cor.pvals){
  xrange <- range(-log(res_table$sign.p)) # set x-axis range
  yrange <- range(res_table$unique.genepairs) # set y-axis range
  plot(xrange, yrange, type = "n", main=paste("corr pval = ", p), xlab = "-log(sign.p)", ylab = " #unique gene pairs") # empty plot
  colors <- rainbow(length(unique(res_table$nonsign.p)))
  for (i in 1:length(unique(res_table$nonsign.p))){
    x <- -log(res_table[nonsign.p==nonsign.p[i] & cor.p==p]$sign.p)
    y <- res_table[nonsign.p==nonsign.p[i] & cor.p==p]$unique.genepairs
    lines(x,y,  type="l", lwd=1.5,lty=linetype[i], col=colors[i])
  }
}
```


Plot without accounting for corr.pval  
different line colors represent different non-sign pval threshold  
x-axis represents the significant p-val treshold  
```{r}
par(mfrow=c(1,1))

xrange <- range(-log(res_table$sign.p)) # set x-axis range
yrange <- range(res_table$unique.genepairs) # set y-axis range
plot(xrange, yrange, type = "n", xlab = "-log(p)", ylab = "#unique gene pairs",  main="#gene pairs where A->B") # empty plot
colors <- rainbow(length(unique(res_table$nonsign.p)))
linetype <- c(1:length(unique(res_table$nonsign.p)))
for (i in 1:length(unique(res_table$nonsign.p))){
  x <- -log(res_table$sign.p[res_table$nonsign.p==unique(res_table$nonsign.p)[i]])
  y <- res_table$unique.genepairs[res_table$nonsign.p==unique(res_table$nonsign.p)[i]]
  lines(x,y,  type="l", lwd=1.5,lty=linetype[i], col=colors[i])
}
legend(x=20, yrange[2], unique(res_table$nonsign.p), col=colors, lty=linetype, cex=0.8)
```

## get table with number of times each geneA appears  
(number of times each geneA is the causal one)
```{r}
ngeneA <- find.effects_TF[find.effects_TF$`A->B`==T, .N, by=geneA]
ngeneA[order(-N)]
```
(number of times each geneB is the affected)
```{r}
ngeneB <- find.effects_TF[find.effects_TF$`B->A`==F, .N, by=geneB]
ngeneB[order(-N)]
```

## combine the two tables to create a table with num of times a gene is causal or is affected
```{r}
ngeneA_B <- merge(ngeneA, ngeneB, by.x = "geneA", by.y = "geneB", all=T)
colnames(ngeneA_B) <- c("gene", "A","B")
ngeneA_B[order(-A)]

#fwrite(ngeneA_B, "/Users/Carolina/Documents/GitHub/DegreeProject/results/2020-01-22/numtimesgenecausal_ornot.gz")
```
### genes that have more connections going into it (that are mostly affected by another gene)
```{r}
ngeneA_B_affected <- ngeneA_B[B>A]
ngeneA_B_affected [order(-B)]
```

### genes that are only causal
```{r}
ngeneA_B_onlycausal <- ngeneA_B[is.na(B)]
ngeneA_B_onlycausal[order(-A)]
```

### genes that are only affected by another gene
```{r}
ngeneA_B_onlyaffected <- ngeneA_B[is.na(A)]
ngeneA_B_onlyaffected[order(-B)]
```


```{r}
plot_YAL003W <- graph_from_edgelist(as.matrix(find.effects_TF[geneA=="YAL003W" | geneB=="YAL003W",.(geneA, geneB)]),directed=TRUE)
```



```{r}
V(plot_YAL003W)$size <- 5
plot(plot_YAL003W, layout=layout_as_tree,edge.arrow.size=.5, vertex.label=NA)
plot(plot_YAL003W, layout=layout_as_tree,edge.arrow.size=.5)
```

# NETWORK ENRICHMENT ANALYSIS
```{r}
library(neat)
`%notin%` <- Negate(`%in%`)
```

  
Some of my genes are not in the kegg or the goslimproc lists  
(genes from both geneA and geneB)
```{r}
print("number of genes in my dataset not in goslimproc")
length(unique(find.effects_TF[which(unique(c(find.effects_TF$geneA, find.effects_TF$geneB)) %notin% unlist(yeast$goslimproc))]$geneA))
print("number of genes in my dataset not in kegg")
length(unique(find.effects_TF[which(unique(c(find.effects_TF$geneA, find.effects_TF$geneB)) %notin% unlist(yeast$kegg))]$geneA))
```

NAB <- number of arrows going from genes in A to genes belonging to B

nodes	-- Vector containing the (ordered) names of all nodes in the network ## what do they mean by ordered??

```{r}
alist <- list(geneA = find.effects_TF$geneA, geneB=find.effects_TF$geneB)
blist <- yeast$kegg

kegg_neat <- data.table(neat(alist = alist, blist = blist , network = as.matrix(find.effects_TF[,.(geneA, geneB)]),nettype = 'directed', nodes = unique(c(find.effects_TF$geneA, find.effects_TF$geneB)), alpha = 0.01, mtc.type = 'fdr', anames=c("genesA", "genesB")))
kegg_neat[order(pvalue)]

blist <- yeast$goslimproc # GOslim process ontology (Ashburner et al., 2000) for the buddying yeast Saccaromyces Cerevisiae
goslimproc_neat <- data.table(neat(alist = alist, blist = blist , network = as.matrix(find.effects_TF[,.(geneA, geneB)]),nettype = 'directed', nodes = unique(c(find.effects_TF$geneA, find.effects_TF$geneB)), alpha = 0.01, mtc.type = 'fdr', anames=c("genesA", "genesB")))
goslimproc_neat[order(pvalue)]
```
using kegg or goslimproc gives different results


```{r}
print("kegg")
table(kegg_neat$conclusion)
print("goslimproc")
table(goslimproc_neat$conclusion)
```

```{r}
over_kegg_neat.over <- data.table(kegg_neat[conclusion=='Overenrichment'])
over_kegg_neat.over[order(pvalue)]

goslimproc_neat.over <- data.table(goslimproc_neat[conclusion=='Overenrichment'])
goslimproc_neat.over[order(pvalue)]
```

```{r}
hist(kegg_neat$pvalue)
hist(kegg_neat$adjusted_p)
hist(goslimproc_neat$pvalue)
hist(goslimproc_neat$adjusted_p)
```


```{r}
over_kegg_neat.under <- data.table(kegg_neat[conclusion=='Underenrichment'])
over_kegg_neat.under

goslimproc_neat.under <- data.table(goslimproc_neat[conclusion=='Underenrichment'])
goslimproc_neat.under
```




## Compare with yeastnet
(data(yeast), yeast$yeastnet)
YeastNet comprises 362512 edges connecting 5808 genes.