---
title: "Exploring chromosome 12"
output:
  html_document:
    fig_caption: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set('python', engine.path = '/usr/local/bin/python3')
knitr::opts_chunk$set(python.reticulate = T)

knitr::opts_knit$set(root.dir = normalizePath("/Users/Carolina/Documents/GitHub/DegreeProject/summarizing/")) 
options(tinytex.verbose = TRUE)
```


# Going to look closer at chromosome 12
```{r message=FALSE}
library("data.table")
library("plotfunctions")
source("myfunctions.R")
```

```{r}
genepos <- fread("results/gene_pos.gz")
if (!exists("find.effects_TF")){
  find.effects_TF <- fread("results/findeffects_TF_newparams.gz")
}

# Add chomosome and start position to each gene
causal.pos.A <- merge(find.effects_TF, genepos, by.x="geneA", by.y="gene", all.x=T)
colnames(causal.pos.A) <- c("geneA", "geneB", "eqtl.A", "eqtl.B", "A->B", "B->A", "strand.A", "start.A","end.A", "chr.A")
causal.pos.B <- merge(causal.pos.A, genepos, by.x="geneB", by.y="gene", all.x=T)


# Keep olnly columns with genes, start positions and chromosomes 
causal.pos.B.2 <- unique(causal.pos.B[,.(geneA, geneB, start.A, chr.A, chr.start, chr.id)])
colnames(causal.pos.B.2) <- c("geneA", "geneB", "start.A", "chr.A", "start.B", "chr.B")

## transform chromosome ids into numbers
# remove "chr" part of the chromosome name
causal.pos.B.2$chr.A <-gsub('chr', '', causal.pos.B.2$chr.A)
causal.pos.B.2$chr.B <-gsub('chr', '', causal.pos.B.2$chr.B)
colnames(causal.pos.B.2) <- c("geneA", "geneB", "start.A", "chr.A", "start.B", "chr.B")

# convert roman chromosome numbers to numbers
causal.pos.B.2$chr.A <- as.numeric(as.roman(causal.pos.B.2$chr.A))
causal.pos.B.2$chr.B <- as.numeric(as.roman(causal.pos.B.2$chr.B))

# order values
causal.pos.B.2.order <- causal.pos.B.2[order(chr.A, start.A, chr.B, start.B)]


# organize coordinates so that they are ordered by chromosome
# vector of chromosomes
vchr <- 1:16
# how much space will be separating chromosomes
separator <- 1e5

coordinates_plot <- sort_by_chr(vchr = vchr, causal.pos.B.2.order, separator = separator)
```

```{r}
if (!exists("find.effects")){
  find.effects <- fread("results/findeffects_all_newparams.gz")
}
#Create a function to generate a continuous color palette
rbPal <- colorRampPalette(c('blue','red'))


coordinates_plot_cor <- merge(coordinates_plot, find.effects[,.(geneA, geneB, cor)], by=c("geneA", "geneB"), all.x=T)
coordinates_plot_cor$cor <- abs(coordinates_plot_cor$cor)
coordinates_plot_cor <- coordinates_plot_cor[order(cor)]
coordinates_plot_cor$col <- rbPal(100)[as.numeric(cut(coordinates_plot_cor$cor,breaks = 100))]

plot_sorted_coordinates(coordinates_plot_cor, separator = separator, col = coordinates_plot_cor$col)
gradientLegend(format(round(coordinates_plot_cor$cor, 3), nsmall = 3), rbPal(100), inside=F, side=3)
```

# Focus on causal gene 12
```{r warning=FALSE}
# coordinates_plot_cor[chr.A==12]
plot_sorted_coordinates(coordinates_plot_cor[chr.A==12], separator = separator, col = coordinates_plot_cor[chr.A==12]$col)
```

# Focus on the group of vertical bands
Chromosome 3 seems to mostly be affected by the genes in the vertical bands so I'm going to look closer at chromosome 3
```{r warning=FALSE}
# the first gene of chromosome 12 that affects chromosome 3
lower_limit <- coordinates_plot_cor[chr.A==12 & chr.B==3][order(start.A)][1]$start.A
# the last gene on chromosome 12 that affects chromosome 3 and that's in the band (there's an extra gene further away from the band)
top_limit <- coordinates_plot_cor[chr.A==12 & chr.B==3][order(start.A)][nrow(coordinates_plot_cor[chr.A==12 & chr.B==3])-1]$start.A

# plot chromosome 12 - the green lines are the limits of the group of vertical bands
plot_sorted_coordinates(coordinates_plot_cor[chr.A==12], separator = separator, col = coordinates_plot_cor[chr.A==12]$col)
abline(v = c(lower_limit-1500, top_limit+1500), col="green")
```


there are `r nrow(unique(coordinates_plot_cor[chr.A==12 & chr.B==3][order(start.A)][,.(geneA, start.A)])[between(start.A,lower_limit, top_limit )])` genes between the two green bands

### get gene names for genes on chromosome 12
```{r}
if (!exists("genes_GO.bio")){
  genes_GO.table <- fread("results/genelistwithGOterm.txt")
  genes_GO.table <- unique(genes_GO.table)
  genes_GO.bio   <- unique(genes_GO.table[GO.namespace=="biological_process"])
}
```

```{r}
genesA_start_order <- unique(coordinates_plot_cor[chr.A == 12 & chr.B == 3][order(start.A)][, .(geneA, start.A)])
chr12_genenames_chr3 <- merge(genesA_start_order, unique(genes_GO.bio[,.(gene, symbol, gene.name, GO.term)]), by.x="geneA", by.y="gene")
chr12_genenames <- merge(unique(coordinates_plot_cor[chr.A == 12][order(start.A)][, .(geneA, start.A)]), unique(genes_GO.bio[,.(gene, symbol, gene.name, GO.term)]), by.x="geneA", by.y="gene")
```

## Look at the first group
```{r warning=FALSE}
# genes that are affecting chromosome 3 + positions
band1.left <- min(genesA_start_order$start.A)
band1.right <- 8525907
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A == 12],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col,
  xlim = c(min(genesA_start_order$start.A), top_limit)
)
abline(v=c(band1.left-1500, band1.right+1500), col="green")
```

There are `r nrow(genesA_start_order[between(start.A,7422391-1500, 7425907+1500)])` between the green lines

```{r warning=FALSE}
# genes that are affecting chromosome 3 + positions
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A == 12],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col,
  xlim=c(band1.left, band1.right)
)
abline(v=c(band1.left-100, band1.right+100), col="green")
```

### genes on the first "band" of chromosome 12
```{r}
# gene names
unique(chr12_genenames[between(start.A,band1.left, band1.right)]$gene.name)
# GO terms present
unique(chr12_genenames[between(start.A,band1.left-1500, band1.right+1500)]$GO.term)
```

## Look at the second group
```{r warning=FALSE}
band2.left <- 8563067
band2.right <- 8626178
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A == 12],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col,
  xlim = c(band2.left, band2.right)
)
abline(v=c(band2.left-1500, band2.right+1500), col="green")
```


There are `r nrow(genesA_start_order[between(start.A,7463067-1500, 7526178+1500)])` genes between the green lines

#### Looking closer
```{r warning=FALSE}
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A == 12],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col,
  xlim = c(band2.left, band2.right)
)
abline(v=c(band2.left-1500, band2.right+1500), col="green")
```

#### genes on the second band of chromsome 12
```{r}
# gene names
unique(chr12_genenames[between(start.A,band2.left-1500, band2.right+1500)]$gene.name)
# GO terms present
unique(chr12_genenames[between(start.A,band2.left-1500, band2.right+1500)]$GO.term)
```
## Look at the last group

```{r warning=FALSE}
band3.left <- 8641374
band3.right <- 8684887
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A == 12 & chr.B==3],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col,
  xlim = c(band3.left, 8844476)
)
abline(v=c(band3.left-1500, band3.right+1500), col="green")
```


There's one extra gene that does not belong to the "band"

```{r warning=FALSE}
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A == 12],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col,
  xlim = c(band3.left, band3.right)
)
abline(v=c(band3.left-1500, band3.right+1500), col="green")
```


#### genes on the last band of chromsome 12
```{r}
# gene names
unique(chr12_genenames[between(start.A,band3.left-1500, band3.right+1500)]$gene.name)
# GO terms present
unique(chr12_genenames[between(start.A,band3.left-1500, band3.right+1500)]$GO.term)
```

# checking if some of my genes in chromosome 12 are inside the described hotspots
```{r}
hotspot_data <- fread("data/SI_Data_08_hotspotTableForPaper_topGenes100_170420_page1.csv")
```

```{r}
hotspot_data.interval <- hotspot_data[,.(hotspotMarker, chromosome, bootstrapIntervalLeft, bootstrapIntervalRight)]
# remove "chr" part of the chromosome name
hotspot_data.interval$chromosome <-gsub('chr', '', hotspot_data.interval$chromosome)

# convert roman chromosome numbers to numbers
hotspot_data.interval$chromosome <- as.numeric(as.roman(hotspot_data.interval$chromosome))

genes_inside_hotspot(hotspot_data = hotspot_data.interval,causal.pos.B.2.order, 12)
```

```{r}
hotspot_data.interval.12 <- hotspot_data.interval[chromosome==12]

bothspositions <- merge(causal.pos.B.2.order, coordinates_plot_cor, by=c("geneA", "geneB", "chr.A", "chr.B"))
setnames(bothspositions, c("geneA", "geneB", "chr.A", "chr.B", "ori.start.A", "ori.start.B", "map.start.A", "map.start.B", "cor","col"))

bothspositions12 <- unique(bothspositions[chr.A==12][,.(geneA, chr.A, ori.start.A, map.start.A)])
```

```{r warning=FALSE}
plot_sorted_coordinates(
  coordinates_plot_cor[chr.A==12],
  separator = separator,
  col = coordinates_plot_cor[chr.A==12]$col
)
abline(v = c(7486306-100, 7491601+100),col="green")
```

